---
format:
  gfm:
    toc: false
    echo: false
execute:
  echo: false
  warning: false
  message: false
---
```{r setup, include = F}
source("code/_master_code.R")
```
# Distribuzione Von Mises

## Introduzione 

Il post di [Daniel Whittenbury](https://dlwhittenbury.github.io/ds-1-sampling-and-visualising-the-von-mises-distribution.html) contiene alcune note riguardanti la distribuzione von Mises.
 
La "statistica direzionale" ha numerosi usi nelle scienze
  
-   La statistica classica, come sviluppata da Gauss, assume che gli errori di misura siano "piccoli" e distribuiti in uno **spazio lineare infinito** (come la retta dei numeri reali o uno spazio euclideo)
-   Questo funziona bene quando le misure sono **molto precise** come nel caso di astronomia o dei geometria che usano strumenti accurati
-   **Le misure angolari** (come direzioni su una bussola, ore del giorno, orientamenti spaziali, ecc.) **vivono naturalmente su una struttura topologica diversa** (un cerchio o una sfera). 
  -   Se gli errori diventano grandi (cioè, se i dati sono molto dispersi), **non ha più senso ignorare la curvatura** dello spazio su cui si trovano i dati.
  -   Per esempio, dire che la media tra 10° e 350° è 180° è assurdo in un contesto circolare, ma potrebbe sembrare corretto se si ignorasse la topologia.
 
La distribuzione di von Mises è stata introdotta da Richard von Mises nel 1918 e la sua PDF è 
  \begin{align} 
  f(x | \mu, \kappa) =  \frac{\exp(k \cos(x - \mu))}{2 \pi I_0(k)}
  \end{align}
 
  - x è la variabile aleatoria angolare 
  - $\mu$ è la **direzione media** (con $\mu \in [0, 2 \pi)$)
  - $\kappa$ è un parametro di concentrazione della distribuzione, che può assumere valori non negativi ($\kappa \in [0, \infty)$)
      - più è grande $\kappa$, più la distribuzione è concentrata attorno alla direzione media
  - $I_0(\kappa)$ è la funzione di Bessel modificata del primo tipo e ordine zero 


Le distribuzioni circolari hanno diversi vantaggi 
 
  - sono definite su uno spazio campionario corretto 
      - **la normale può produrre stime errate vicino ai limiti (es. 0 e $2\pi$)**
  - la distribuzione di von Mises ha un comportamento limite corretto quando $\kappa \to 0$: tende a una distribuzione uniforme sul cerchio dove tutte le direzioni sono equiprobabili
  - **usare la normale in problemi circolari può portare a errori seri come fondere in modo scorretto una distribuzione a priori con una funzione di verosimiglianza**
      - **nelle misurazioni dell’angolo di arrivo (in applicazioni di tracciamento), si rischia di aggiornare male la stima bayesiana se si ignora la natura ciclica della variabile**
      
## Optim function
  
A common task in statistics is maximizing (or minimizing) complex univariate or multivariate functions. This is typically done in the context of maximizing likelihood functions with respect to a vector of unknown parameters, given the observed data.
  
-   More generally, this task pertains to an “optimization” problem, where we seek to finding a set of parameter values that minimizes or maximizes some pre-defined objective function. 
  - When this objective function is a log likelihood function, this reduces to the problem of maximum likelihood estimation.
  - There are many potential approaches to use for optimization in this context. 
  - How best to optimize a function of interest depends on the nature of the function to be optimized, as well as practical concerns regarding a candidate procedure to be used. 
  
By default `optim` performs minimization, but it will maximize if control$fnscale is negative. optimHess is an auxiliary function to compute the Hessian at a later stage if hessian = TRUE was forgotten.
  
-   The default method is an implementation of that of Nelder and Mead (1965), that uses only function values and is robust but relatively slow. It will work reasonably well for non-differentiable functions.
-   Method "BFGS" is a quasi-Newton method (also known as a variable metric algorithm), specifically that published simultaneously in 1970 by Broyden, Fletcher, Goldfarb and Shanno. This uses function values and gradients to build up a picture of the surface to be optimized.
-   Method "CG" is a conjugate gradients method based on that by Fletcher and Reeves (1964) (but with the option of Polak–Ribiere or Beale–Sorenson updates). Conjugate gradient methods will generally be more fragile than the BFGS method, but as they do not store a matrix they may be successful in much larger optimization problems.
-   Method "L-BFGS-B" is that of Byrd et al. (1995) which allows box constraints, that is each variable can be given a lower and/or upper bound. The initial value must satisfy the constraints. This uses a limited-memory modification of the BFGS quasi-Newton method. If non-trivial bounds are supplied, this method will be selected, with a warning.
-   Nocedal and Wright (1999) is a comprehensive reference for the previous three methods.
-   Method "SANN" is by default a variant of simulated annealing given in Belisle (1992). 
  -   Simulated-annealing belongs to the class of stochastic global optimization methods. It uses only function values but is relatively slow. 
  -   It will also work for non-differentiable functions. This implementation uses the Metropolis function for the acceptance probability. 
  -   By default the next candidate point is generated from a Gaussian Markov kernel with scale proportional to the actual temperature. If a function to generate a new candidate point is given, method "SANN" can also be used to solve combinatorial optimization problems. 
    -   Temperatures are decreased according to the logarithmic cooling schedule as given in Belisle (1992, p. 890); specifically, the temperature is set to ` temp / log(((t-1) %/% tmax) * tmax + exp(1))`, where `t` is the current iteration step and `temp` and `tmax` are specifiable via control, see below. 
    -   Note that the "SANN" method depends critically on the settings of the control parameters. It is not a general-purpose method but can be very useful in getting to a good value on a very rough surface.
    -   Method "Brent" is for one-dimensional problems only, using `optimize(<ff>, lower, upper, tol = control$reltol)` where `<ff>` is `function(par) fn(par, ...)/control$fnscale`. It can be useful in cases when `optim()` is used inside other functions where only method can be specified, such as in `mle` from package stats4.
    
Function `fn` can return `NA` or `Inf` if the function cannot be evaluated at the supplied value, but the initial value must have a computable finite value of fn. (Except for method "L-BFGS-B" where the values should always be finite.)
  
`optim` can be used recursively, and for a single parameter as well as many. It also accepts a zero-length par, and just evaluates the function with that argument

## Stima di una Von Mises con una covariata
 
Creo due vettori $\mathbf{x}_n$ e $\mathbf{y}_n$ in cui 

 -  $X_i \sim N(0, 1)$
 -  $Y_i \sim VonMises(\mu = 2atan(\beta_0 +\beta_1 \cdot x_i), k = 10)$ con $\beta_0=-1$, $\beta_1 = 3.5$

```{r}
#| label: random-data
rndm_x <- rnorm(1000)
rndm_y <- sapply(rndm_x, function(xi) circular::rvonmises(n = 1, mu = 2*atan(-1 + 3.5 * xi), kappa = 10))
```
  
Definisco la funzione di - log-verosimiglianza

```{r}
#| label: log-lik-function
log.lik.vm <- function(par, x, y){
  beta.0 <- par[1]
  beta.1 <- par[2]
  kappa <- exp(par[3])
  
  l <- array(dim = length(x))
  
  for(i in seq_along(x)){
    l[i] = kappa * cos(y[i] - 2 * atan(beta.0 + beta.1*x[i])) - log((besselI(kappa, nu = 0)))
  }
  
  return(-sum(l))
}
```

Minimizzo la funione con `optim`

```{r}
#| label: optim-results
result <- optim(
  par = c(beta.0 = 0, beta.1 = 0, kappa = log(2)),
  fn = log.lik.vm,
  x = rndm_x,
  y = rndm_y, 
  hessian = T)

# Standard error
det(result$hessian) != 0 # condizione di invertibilità
cov.matrix <- solve(result$hessian)
se <- sqrt(diag(cov.matrix))

# Intervalli di confidenza asintotici (95%)
tibble(
  parametri = c("beta 0", "beta 1", "kappa"), 
  estimate = result$par,
  se = se,
  lower = estimate - qnorm(0.975) * se,
  upper = estimate + qnorm(0.975) * se
) %>% 
  kable(
    col.names = c("Parameter", "Estimate", "Std. Error", "95% CI Lower", "95% CI Upper"),
    align = "lcccc",
    format = "markdown", 
    digits = 3
  )
```

Nota esplicativa: 
 
 - Il parametro `kappa` è stimato in scala logaritmica per garantire che il valore restituito sia sempre positivo. Il valore `par[3]` rappresenta quindi il logaritmo naturale del parametro reale, e viene esponenziato all'interno della funzione di verosimiglianza.
 - Impostando `hessian = TRUE`, `optim()` restituisce anche la **matrice Hessiana** della funzione obiettivo valutata nel punto di minimo, che corrisponde (sotto opportune condizioni regolarità) alla **matrice di informazione osservata**.
 - Se questa matrice Hessiana è invertibile, la sua inversa fornisce una **stima asintotica della matrice di varianza-covarianza** dei parametri stimati.
 - Le **radici quadrate degli elementi diagonali** della matrice inversa corrispondono agli **errori standard asintotici** delle stime dei parametri.
 
## Simulazione di $K$ repliche per lo stesso set di parametri veri
 
Genero $K$ set di dati random e stimo i parametri che massimizzano la verosimiglianza per ogni set di dati 

```{r}
#| label: k-data-set
gen.data <- function(n, beta.0, beta.1, kappa){
  x <- rnorm(n)
  y <- sapply(x, function(xi) circular::rvonmises(n= 1, mu= 2*atan(beta.0 + beta.1 * xi), kappa = kappa))
  data <- tibble(x=x, y=y)
}
```

Per riproducibilità genero una sola volta una lista di $K = 1000$ `tibble` inserendo come come parametri `n = 1000`, `beta.0 = -1`, `beta.1 = 3.5`, `kappa = 10` e la salvo in un file chiamato `sim_1.RData`

```{r k-simulations}
load(paste0(dir_data, "/sim_1.RData"))

results <- map_dfr(
  .x = sim_1,
  .f = ~ {
    mle <- optim(
      par = c(beta.0 = 0, beta.1 = 0, kappa = log(2)),
      fn = log.lik.vm,
      x = .x$x,
      y = .x$y
    )
    
    tibble(
      beta.0 = mle$par[1],
      beta.1 = mle$par[2],
      kappa  = exp(mle$par[3]),
      convergence = mle$convergence
    )
  }
)

# Visualizzazione ----
summary_df <- tibble(
  Parameter = c("beta 0", "beta 1", "kappa"),
  Mean      = c(mean(results$beta.0), mean(results$beta.1), mean(results$kappa)),
  `Std. Dev` = c(sd(results$beta.0), sd(results$beta.1), sd(results$kappa)),
  `2.5%`    = c(quantile(results$beta.0, 0.025),
                quantile(results$beta.1, 0.025),
                quantile(results$kappa, 0.025)),
  `97.5%`   = c(quantile(results$beta.0, 0.975),
                quantile(results$beta.1, 0.975),
                quantile(results$kappa, 0.975))
)

summary_df %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(
    align = "lcccc",
    col.names = c("Parameter", "Mean", "Std. Dev", "2.5%", "97.5%"),
    format = "markdown"
  )
```